{\bf Practical Concurrency Testing} \\ \normalsize \vspace{1em} {\em or, How I Learned to Stop Worrying and Love the Exponential Explosion}

### 2017-01-03 proposal/quicksand.tex
quicksand verif

In this chapter, I will present a new algorithm for model checkers to automatically choose which preemption points to use to test reasonably-sized state spaces.
This is the first of the three projects I am proposing for this thesis, and has largely been completed and is under submission.

\section{Motivation}

This project is motivated by my experience contributing to Parrot \cite{parrot},
a partially-determinizing runtime for multithreaded programs which allowed some nondeterminism in performance-critical sections to achieve high performance.
We combined Parrot with dBug \cite{dbug-ssv}, a model checker, to check the correctness of the nondeterministic sections of programs which remained.
Unfortunately, dBug was able to finish a state space for fewer than half of these programs in our evaluation in a 24-hour time limit,
even after the reduction Parrot provided.
In that project, our model checker was limited to a fixed set of preemption points;
in other words, it would always test different programs with the same granularity of thread interleavings.
Consequently, the resulting completion time for each test would vary wildly,
unpredictably falling on one side or the other of our arbitrary time limit.

@@

\begin{algorithm}[t]
        \SetKwInOut{Input}{Input}
        %\textbf{Function} GetBestJob($j_0$, PendingJobs, SuspendedJobs): \\
        \Input{$j_0$, the currently-running job}
        %\Input{$eta$, $j_0$'s predicted completion time}
        \Input{\PendingJobs, the list of pending jobs, sorted by decreasing heuristic priority}
        \Input{\SuspendedJobs, the list of already-suspended jobs, sorted by increasing ETA}
        \Input{$T$, the remaining time in the CPU budget}
        \If{\GetETA{$j_0$} $<$ HeuristicETAFactor $\times$ $T$}{
                return $j_0$ // Common case: job is expected to finish.
        }
        \ForEach{job $j_P \in$ \PendingJobs}{
                // Don't run a pending job if a subset of it is already suspended; its ETA would be at least as bad. \\
                \If {$\forall j_S \in$ \SuspendedJobs, \GetPPSet{$j_S$} $\not\subset$ \GetPPSet{$j_P$}}{
                        return $j_P$
                }
        }
        %// no pending jobs; maybe resume a suspended job \\
        \ForEach{job $j_S \in$ \SuspendedJobs}{
                \If{\GetPPSet{$j_0$} $\not\subset$ \GetPPSet{$j_S$}
                        $\land$
                        \GetETA{$j_0$} $>$ \GetETA{$j_S$}}{
                        // If a subset of $j_S$ is also suspended, don't run the larger one first. \\
                        \If{$\forall j_{S2} \in$ \SuspendedJobs, \GetPPSet{$j_{S2}$} $\not\subset$ \GetPPSet{$j_S$}}{
                                return $j_S$
                        }
                }
        }
        return $j_0$ // \GetETA{$j_0$} was bad, but no other $j$ was better.
        \caption{Suspending exploration of a state space in favour of a potentially smaller one.}
        \label{alg:shouldworkblock}
\end{algorithm}

\newcommand\AllJobs{\ensuremath{\mathcal{J}}}
\begin{algorithm}[t]
        \SetKwInOut{Input}{Input}
        \Input{$j_0$, the currently-running job}
        \Input{\AllJobs, the set of all existing (or completed) jobs}
        \Input{$\alpha$, an instruction reported by the MC as part of a racing access pair}
        \If{$\forall j \in \AllJobs,$
        \GetPPSet{$j_0$} $\cup$ $\alpha$
        $\not\subseteq$
        \GetPPSet{$j$}
        }{
                AddNewJob(\GetPPSet{$j_0$} $\cup$ $\alpha$, HeuristicPriority($\alpha$)) \\
        }
        \If{$\forall j \in \AllJobs,$ \GetPPSet{$j$} $\neq \{yield, \alpha\}$}{
                AddNewJob($\{yield, \alpha\}$, HeuristicPriority($\alpha$))
        }
        \caption{Adding new jobs with data-race PPs.}
        \label{alg:handledatarace}
\end{algorithm}

@@

>>>>

In this chapter, I will present a new algorithm for model checkers to automatically choose which preemption points to use to test reasonably-sized state spaces.
This is the first of the three projects I am proposing for this thesis, and has already been published at OOPSLA 2016 \cite{quicksand}.
The thesis will largely restate the paper, and also go into more detail on some ideas which didn't make it into the paper, such as measuring the variance in each experimental data series.
Here in the proposal I will simply summarize the content of the paper.

\section{Motivation}
\label{sec:qs-motivation}

This project is motivated by my experience contributing to Parrot \cite{parrot},
a (partially) determinizing runtime for multithreaded programs which allowed some nondeterminism in performance-critical sections to achieve high performance.
We combined Parrot with dBug \cite{dbug-ssv}, a model checker, to check the correctness of the nondeterministic sections of programs which remained.
Unfortunately, dBug was able to finish a state space for fewer than half of these programs in our evaluation in a 24-hour time limit,
even after the reduction Parrot provided.
In that project, our model checker was limited to a fixed set of preemption points;
in other words, it would always test different programs with the same granularity of thread interleavings.
Consequently, the resulting completion time for each test would vary wildly,
unpredictably falling on one side or the other of our arbitrary time limit.

@@

\begin{algorithm}[h]
        \SetKwInOut{Input}{Input}
        %\textbf{Function} GetBestJob($j_0$, PendingJobs, SuspendedJobs): \\
        \Input{$j_0$, the currently-running job}
        %\Input{$eta$, $j_0$'s predicted completion time}
        \Input{\PendingJobs, the list of pending jobs, sorted by decreasing heuristic priority}
        \Input{\SuspendedJobs, the list of already-suspended jobs, sorted by increasing ETA}
        \Input{$T$, the remaining time in the CPU budget}
        \If{\GetETA{$j_0$} $<$ HeuristicETAFactor $\times$ $T$}{
                return $j_0$ // Common case: job is expected to finish.
        }
        \ForEach{job $j_P \in$ \PendingJobs}{
                // Don't run a pending job if a subset of it is already suspended; its ETA would be at least as bad. \\
                \If {$\forall j_S \in$ \SuspendedJobs, \GetPPSet{$j_S$} $\not\subset$ \GetPPSet{$j_P$}}{
                        return $j_P$
                }
        }
        %// no pending jobs; maybe resume a suspended job \\
        \ForEach{job $j_S \in$ \SuspendedJobs}{
                \If{\GetPPSet{$j_0$} $\not\subset$ \GetPPSet{$j_S$}
                        $\land$
                        \GetETA{$j_0$} $>$ \GetETA{$j_S$}}{
                        // If a subset of $j_S$ is also suspended, don't run the larger one first. \\
                        \If{$\forall j_{S2} \in$ \SuspendedJobs, \GetPPSet{$j_{S2}$} $\not\subset$ \GetPPSet{$j_S$}}{
                                return $j_S$
                        }
                }
        }
        return $j_0$ // \GetETA{$j_0$} was bad, but no other $j$ was better.
        \caption{Suspending exploration of a state space in favour of a potentially smaller one.}
        \label{alg:shouldworkblock}
\end{algorithm}

\newcommand\AllJobs{\ensuremath{\mathcal{J}}}
\begin{algorithm}[h]
        \SetKwInOut{Input}{Input}
        \Input{$j_0$, the currently-running job}
        \Input{\AllJobs, the set of all existing (or completed) jobs}
        \Input{$\alpha$, an instruction reported by the MC as part of a racing access pair}
        \If{$\forall j \in \AllJobs,$
        \GetPPSet{$j_0$} $\cup$ $\alpha$
        $\not\subseteq$
        \GetPPSet{$j$}
        }{
                AddNewJob(\GetPPSet{$j_0$} $\cup$ $\alpha$, HeuristicPriority($\alpha$)) \\
        }
        \If{$\forall j \in \AllJobs,$ \GetPPSet{$j$} $\neq \{yield, \alpha\}$}{
                AddNewJob($\{yield, \alpha\}$, HeuristicPriority($\alpha$))
        }
        \caption{Adding new jobs with data-race PPs.}
        \label{alg:handledatarace}
\end{algorithm}

@@

\subsection{Convergence of Iterative Deepening}
When the a test's state spaces are small enough that Quicksand can exhaustively check all of them within a given CPU budget,
the resulting verification claim turns out to be equivalent to a na\"{i}model checker which preempts on every memory access.
In this way, the Iterative Deepening algorithm provides the ``best of both worlds'' for the tradeoff mentioned in Section~\ref{sec:qs-motivation}:
when tests are too large, Quicksand falls back on prioritization heuristics to find bugs quickly;
when tests are tractable, Quicksand's verification is equal to the strongest from any single-state-space technique.

The theorem statement is as follows:

\begin{theorem}[Convergence]
	If a bug can be exposed by any
thread interleaving possible by preempting on all instructions during a specific test, Iterative Deepening will eventually test an equivalent interleaving which exposes the same bug.
\end{theorem}

The proof is available at \cite{quicksand-soundness}.

\subsection{Pruning false-positive data-race candidates}
I have also proved a (somewhat less remarkable) theorem concerning the soundness of a new tactic for pruning certain types of false-positive data race candidates (under Limited HB).
A ``malloc-recycle'' false positive occurs when two threads seem to race on the same memory address, but only because the {\tt malloc}ed block containing that address was {\tt free}d and reallocated in between.
If the transitions of the two threads were reordered, the address collision would disappear and there would be no race.
The proof is available at \cite{quicksand-soundness}; the theorem statement is:

\begin{theorem}[Soundness of eliminating malloc-recycle candidates]
	If a malloc-recycle candidate is not a false positive,
	DPOR will test an alternate thread interleaving in which the
	accesses can race without fitting the malloc-recycle pattern.
\end{theorem}
